Mixing Time Estimation in Reversible Markov Chains from a Single Sample Path

### 2015-10-31 arxiv-v2/appendix-empirical.tex
corner cases
freedman

We make use of a simple corollary of Freedman's inequality for
martingales~\citep[Theorem 1.6]{Fre75}.
\begin{theorem}[Freedman's inequality]
  \label{thm:freedman}
  Let $Y_1, Y_2, \dotsc, Y_n$ be a bounded martingale difference
  sequence with respect to the filtration $\cF = (\cF_i)_{1 \leq i
  \leq n}$ and with $|Y_i| \leq b$ for some $b>0$.
  Let $V := \sum_{i=1}^n \EE{Y_i^2|\cF_{i-1}}$ and $S := \max_{1 \leq
  m \leq n} \sum_{i=1}^m Y_i$.
  For all $t, v > 0$,
  \[
    \Pr\Brackets{
      S > \sqrt{2vt} + 2bt/3
      \,\wedge\,
      V \leq v
    }
    \leq e^{-t}
    .
  \]
\end{theorem}
\begin{corollary}
  \label{cor:freedman}
  Under the same setting as \Cref{thm:freedman}, for any $t > 0$ and
  $c > 1$,
  \[
    \Pr\Brackets{
      S > \sqrt{2cVt} + 5bt/3
    }
    \leq
    \Parens{ 1 + \ceil{\log_c(2n/t)} }
    e^{-t}
    .
  \]
\end{corollary}
\begin{proof}
  Define $v_i := c^i b^2t/2$ for $i = 0, 1, 2, \dotsc, \ceil{
  \log_c(2n/t) }$, and let $v_{-1} := -\infty$.
  Then, since $V \in [0,b^2n]$,
  \begin{align*}
    \Pr\Brackets{
      S > \sqrt{2\max\braces{v_0,cV}t} + 2bt/3
    }
    & =
    \sum_{i=0}^{\ceil{\log_c(2n/t)}}
    \Pr\Brackets{
      S > \sqrt{2\max\braces{v_0,cV}t} + 2bt/3
      \,\wedge\, v_{j-1} < V \leq v_j
    }
    \\
    & \leq
    \sum_{i=0}^{\ceil{\log_c(2n/t)}}
    \Pr\Brackets{
      S > \sqrt{2\max\braces{v_0,cv_{j-1}}t} + 2bt/3
      \,\wedge\, v_{j-1} < V \leq v_j
    }
    \\
    & \leq
    \sum_{i=0}^{\ceil{\log_c(2n/t)}}
    \Pr\Brackets{
      S > \sqrt{2v_jt} + 2bt/3
      \,\wedge\, V \leq v_j
    }
    \\
    & \leq
    \Parens{ 1 + \ceil{\log_c(2n/t)} }
    e^{-t}
  \end{align*}
  where the final inequality uses \Cref{thm:freedman}.
  The conclusion now follows because
  \[
    \sqrt{2cVt} + 5bt/3
    \geq \sqrt{2\max\braces{v_0,cV}t} + 2bt/3
    .
    \qedhere
  \]
\end{proof}

\begin{lemma}
  \label{lem:P-unobs-bound}
  The following holds for any constant $c>1$ with probability at least
  $1-\delta$: for all $(i,j) \in \iset{d}^2$,
  \begin{equation}
    \abs{ \wh{P}_{i,j} - P_{i,j} }
    \leq
    \sqrt{\Parens{\frac{N_i}{N_i+d\alpha}}\frac{2cP_{i,j}(1-P_{i,j})\emptail}{N_i+d\alpha}}
    + \frac{(5/3)\emptail}{N_i+d\alpha}
    + \frac{\abs{\alpha-d\alpha P_{i,j}}}{N_i+d\alpha}
    \label{eq:P-unobs-bound}
  \end{equation}
  where
  \begin{equation}
    \emptail
    := \inf
    \Braces{
      t\geq0 :
      2d^2 \Parens{ 1 + \ceil{\log_c(2n/t)} } e^{-t} \leq \delta
    }
    = O\Parens{ \log\Parens{ \frac{d\log(n)}{\delta} } }
    \,.
    \label{eq:emptail}
  \end{equation}
\end{lemma}
\begin{proof}
  For each $t \in \bbN$, define $\cF_t$ to be the $\sigma$-field
  generated by $X_1,X_2,\dotsc,X_t$.
  Fix a pair $(i,j) \in [d]^2$, and define
  \[
    Y_{t+1} := \Ind{X_t = i} (\Ind{X_{t+1}=j} - P_{i,j})
    , \quad t \in [n-1]
  \]
  so that
  \[
    \sum_{t=1}^{n-1} Y_{t+1}
    = N_{i,j} - N_i P_{i,j}
    .
  \]
  By the Markov property, the stochastic process $\{ Y_{t+1} \}_{t \in
  [n-1]}$ is an $\cF_t$-adapted martingale difference sequence:
  $Y_{t+1}$ is $\cF_{t+1}$-measurable and $\EE{Y_{t+1}| \cF_t } = 0$,
  for each $t \in [n-1]$.
  Moreover, for each $t \in [n-1]$,
  \[
    \EE{Y_{t+1}^2| \cF_t } = \Ind{X_t=i} P_{i,j} (1-P_{i,j})
  \]
  and
  \[
    Y_{t+1} \in [-P_{i,j},1-P_{i,j}] .
  \]
  Therefore, by \Cref{cor:freedman} and union bounds, we have
  \[
    \abs{ N_{i,j} - N_i P_{i,j} }
    \leq \sqrt{2c N_i P_{i,j} (1-P_{i,j}) \emptail} + \frac53
    \emptail
  \]
  for all $(i,j) \in [d]^2$.
\end{proof}

>>>>

We make use of a simple corollary of Freedman's inequality for
martingales~\citep[Theorem 1.6]{Fre75}.
\begin{theorem}[Freedman's inequality]
  \label{thm:freedman}
  Let $(Y_i)_{i \in \bbN}$ be a bounded martingale difference sequence
  with respect to the filtration $(\cF_i)_{i \in \bbN}$; assume for
  some $b>0$, $|Y_i| \leq b$ almost surely for all $i \in \bbN$.
  Let $V_k := \sum_{i=1}^k \EE{Y_i^2|\cF_{i-1}}$ and $S_k :=
  \sum_{i=1}^k Y_i$ for $k \in \bbN$.
  For all $t, v > 0$,
  \[
    \Pr\Brackets{
      \exists k \in \bbN \,\centerdot\,
      S_k > \sqrt{2vt} + 2bt/3
      \,\wedge\,
      V_k \leq v
    }
    \leq e^{-t}
    .
  \]
\end{theorem}
\begin{corollary}
  \label{cor:freedman}
  Under the same setting as \Cref{thm:freedman}, for any $n \geq 1$,
  $t > 0$, and $c > 1$,
  \[
    \Pr\Brackets{
      \exists k \in [n] \,\centerdot\,
      S_k > \sqrt{2cV_kt} + 5bt/3
    }
    \leq
    \Parens{ 1 + \ceil{\log_c(2n/t)}_+ }
    e^{-t}
    .
  \]
\end{corollary}
\begin{proof}
  Define $v_i := c^i b^2t/2$ for $i = 0, 1, 2, \dotsc, \ceil{
  \log_c(2n/t) }_+$, and let $v_{-1} := -\infty$.
  Then, since $V_k \in [0,b^2n]$ for all $k\in[n]$,
  \begin{align*}
    \lefteqn{
      \Pr\Brackets{
        \exists k \in [n] \,\centerdot\,
        S_k > \sqrt{2\max\braces{v_0,cV_k}t} + 2bt/3
      }
    }
    \\
    & =
    \sum_{i=0}^{\ceil{\log_c(2n/t)}_+}
    \Pr\Brackets{
      \exists k \in [n] \,\centerdot\,
      S_k > \sqrt{2\max\braces{v_0,cV_k}t} + 2bt/3
      \,\wedge\, v_{i-1} < V_k \leq v_i
    }
    \\
    & \leq
    \sum_{i=0}^{\ceil{\log_c(2n/t)}_+}
    \Pr\Brackets{
      \exists k \in [n] \,\centerdot\,
      S_k > \sqrt{2\max\braces{v_0,cv_{i-1}}t} + 2bt/3
      \,\wedge\, v_{i-1} < V_k \leq v_i
    }
    \\
    & \leq
    \sum_{i=0}^{\ceil{\log_c(2n/t)}_+}
    \Pr\Brackets{
      \exists k \in [n] \,\centerdot\,
      S_k > \sqrt{2v_it} + 2bt/3
      \,\wedge\, V_k \leq v_i
    }
    \\
    & \leq
    \Parens{ 1 + \ceil{\log_c(2n/t)}_+ }
    e^{-t}\,,
  \end{align*}
  where the final inequality uses \Cref{thm:freedman}.
  The conclusion now follows because
  \[
    \sqrt{2cV_kt} + 5bt/3
    \geq \sqrt{2\max\braces{v_0,cV_k}t} + 2bt/3
  \]
  for all $k \in [n]$.
\end{proof}

\begin{lemma}
  \label{lem:P-unobs-bound}
  The following holds for any constant $c>1$ with probability at least
  $1-\delta$: for all $(i,j) \in \iset{d}^2$,
  \begin{equation}
    \abs{ \wh{P}_{i,j} - P_{i,j} }
    \leq
    \sqrt{\Parens{\frac{N_i}{N_i+d\alpha}}\frac{2cP_{i,j}(1-P_{i,j})\emptail}{N_i+d\alpha}}
    + \frac{(5/3)\emptail}{N_i+d\alpha}
    + \frac{\abs{\alpha-d\alpha P_{i,j}}}{N_i+d\alpha}\,,
    \label{eq:P-unobs-bound}
  \end{equation}
  where
  \begin{equation}
    \emptail
    := \inf
    \Braces{
      t\geq0 :
      2d^2 \Parens{ 1 + \ceil{\log_c(2n/t)}_+ } e^{-t} \leq \delta
    }
    = O\Parens{ \log\Parens{ \frac{d\log(n)}{\delta} } }
    \,.
    \label{eq:emptail}
  \end{equation}
\end{lemma}
\begin{proof}
  For each $t \in \bbN$, define $\cF_t$ to be the $\sigma$-field
  generated by $X_1,X_2,\dotsc,X_t$.
  Fix a pair $(i,j) \in [d]^2$, and define
  \[
    Y_{t+1} := \Ind{X_t = i} (\Ind{X_{t+1}=j} - P_{i,j})
    , \quad t \in [n-1]
  \]
  so that
  \[
    \sum_{t=1}^{n-1} Y_{t+1}
    = N_{i,j} - N_i P_{i,j}
    .
  \]
  By the Markov property, the stochastic process $\{ Y_{t+1} \}_{t \in
  [n-1]}$ is an $\cF_t$-adapted martingale difference sequence:
  $Y_{t+1}$ is $\cF_{t+1}$-measurable and $\EE{Y_{t+1}| \cF_t } = 0$,
  for each $t \in [n-1]$.
  Moreover, for each $t \in [n-1]$,
  \[
    \EE{Y_{t+1}^2| \cF_t } = \Ind{X_t=i} P_{i,j} (1-P_{i,j})
  \]
  and
  \[
    Y_{t+1} \in [-P_{i,j},1-P_{i,j}] .
  \]
  Therefore, by \Cref{cor:freedman} and union bounds, we have
  \[
    \abs{ N_{i,j} - N_i P_{i,j} }
    \leq \sqrt{2c N_i P_{i,j} (1-P_{i,j}) \emptail} + \frac53
    \emptail
  \]
  for all $(i,j) \in [d]^2$.
\end{proof}
